KSPHPDDM and PCHPDDM: extending PETS\MakeLowercase{c} with advanced Krylov methods and robust multilevel Overlapping Schwarz preconditioners

### 2020-07-21 article.tex
Switch to Elsevier template

% SIAM Article Template
\documentclass[review,onefignum,onetabnum]{siamart190516}

% Information that is shared between the article and the supplement
% (title and author information, macros, packages, etc.) goes into
% ex_shared.tex. If there is no supplement, this file can be included
% directly.

\input{shared}

% Optional PDF information
\ifpdf
\hypersetup{
  pdftitle={KSPHPDDM and PCHPDDM: Extending PETSc with Robust Overlapping Schwarz Preconditioners and Advanced Krylov Methods},
  pdfauthor={P. Jolivet, J. E. Roman, and S. Zampini}
}
\fi

% The next statement enables references to information in the
% supplement. See the xr-hyperref package for details.

% \externaldocument{supplement}

% FundRef data to be entered by SIAM
%<funding-group specific-use="FundRef">
%<award-group>
%<funding-source>
%<named-content content-type="funder-name">
%</named-content>
%<named-content content-type="funder-identifier">
%</named-content>
%</funding-source>
%<award-id> </award-id>
%</award-group>
%</funding-group>
\newcommand\col{:}
\newcommand{\pk}[1]{\texttt{#1}}
\usepackage{stmaryrd}
\usepackage{pgfplotstable}
\pgfplotsset{compat=newest}
\pgfplotsset{colormap={paraview}{rgb(0cm)=(0.278431,0.278431,0.858824) rgb(0.1428571429cm)=(0,0,0.360784) rgb(0.2857142857cm)=(0,1,1) rgb(0.4285714286cm)=(0,0.501961,0) rgb(0.5714285714cm)=(1,1,0) rgb(0.7142857143cm)=(1,0.380392,0) rgb(0.8571428571cm)=(0.419608,0,0) rgb(1cm)=(0.878431,0.301961,0.301961)}}
\usepackage{fancyvrb}
\usepackage{subfig}
\usepackage{upquote,textcomp}
\newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
\usepackage{colortbl,booktabs}
\definecolor{last}{RGB}{44,123,182}
\definecolor{blueaccent}{RGB}{215,25,28}
\definecolor{myyellow}{RGB}{102,102,17}
\definecolor{myred}{RGB}{197,65,0}
\definecolor{myblue}{RGB}{0,0,235}
\definecolor{myaqua}{RGB}{9,156,156}
\definecolor{mygreen}{RGB}{63,126,28}
\definecolor{paraviewred}{rgb}{0.278431,0.278431,0.858824}
\definecolor{paraviewblue}{rgb}{0.878431,0.301961,0.301961}
\definecolor{darkgray}{RGB}{253,174,97}
\definecolor{mediumgray}{RGB}{255,255,191}
\usepackage[outline]{contour}
\usepackage{tikz}
\usepackage[binary-units=true]{siunitx}
\usetikzlibrary{decorations.text,arrows.meta,calc,plotmarks,shadows,spy,fixedpointarithmetic,patterns,intersections,arrows.meta,shapes,decorations.text,decorations.pathmorphing,backgrounds,fit,positioning,shapes.symbols,chains,3d,calc,external}
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}
\long\def\ifnodedefined#1#2#3{%
    \@ifundefined{pgf@sh@ns@#1}{#3}{#2}%
}
\makeatother
\usepackage{amsmath,etoolbox}
\AtBeginEnvironment{pmatrix}{\setlength{\arraycolsep}{2pt}}
\pgfmathsetmacro{\dof}{4}
\pgfmathsetmacro{\overlap}{1}
\begin{document}

\maketitle

% REQUIRED
\begin{abstract}
  Contemporary applications in computational science and engineering often
    require the solution of linear systems which may be of different sizes,
    shapes, and structures. The goal of this paper is to explain how two
    libraries, PETSc and HPDDM, have been interfaced together in order to offer
    end-users robust overlapping Schwarz preconditioners and advanced Krylov
    methods featuring recycling and block solution techniques.
    The flexibility of the implementation %, which for example may deal with various matrix formats,
    is showcased and explained with minimalist, easy-to-run, and reproducible examples,
    to ease the integration of these algorithms into more advanced frameworks.
    The examples provided cover applications from
    %simulate the linear or nonlinear solves of typical
    %applications in fracture mechanics, linear stability analysis or more
    %broadly-speaking
    eigenanalysis, elasticity, combustion, and
    electromagnetism.
\end{abstract}

@@

% REQUIRED
\begin{AMS}
  35Q68, 65Y05
\end{AMS}

\section{Introduction}
Computational science and engineering today enjoys unprecedented opportunities to transform
the way society solves many of its most urgent technological problems through predictive simulations.
At the heart of simulations are robust and scalable solution algorithms, and the efficient
production of applications requires a rich ecosystem of state-of-the-art
reusable libraries from which domain
specialists can benefit~\cite{Bartlett:2017:XFT}. In this paper, we focus on the interoperability of two such
libraries.

@@

Fruits of the proposed enhancement are discussed for 
applications from
the Scalable Library for
Eigenvalue Problem computations (SLEPc)~\cite{hernandez2005ssf}, 
an add-on library that extends PETSc
with several classes for linear and nonlinear eigenvalue problems, as well as with
tools to facilitate the use of shift-and-invert spectral transformations.

The paper is divided in two main parts. \Cref{sec:KSPHPDDM} introduces
the interface to recycling and block Krylov methods
(\pk{KSPHPDDM}) and discusses the readiness of the PETSc library with respect to block Krylov methods.
Different applications from eigen-analysis are also provided.
\Cref{sec:PCHPDDM} introduces the suite of robust multilevel overlapping Schwarz methods (\pk{PCHPDDM})
with examples from linear and nonlinear partial differential equations.
Eventually, concluding remarks are provided in \cref{sec:conclusion}.

@@

Because Krylov methods may need long-term recurrences to mitigate round-off
errors during the generation of subspaces, it is common to introduce a
restart parameter in order to control their memory consumption and the volume of
global communications needed when orthonormalizing a candidate basis vector. Such
restarts may hinder convergence of iterative methods, and even introduce
convergence plateaus. Recycling techniques have been introduced to attenuate
these effects. In PETSc, the Loose GMRES~\cite{baker2005technique}~(LGMRES) and the
Deflated GMRES~\cite{erhel1996restarted,wakam2013158}~(DGMRES), implemented as
\pk{KSPLGMRES} and \pk{KSPDGMRES}, are available. However, neither handle variable
preconditioning, and \pk{KSPDGMRES} does not support complex arithmetic;
moreover, extending the recycling capabilites of such methods to sequence of linear systems
with smoothly varying coefficient matrices and right-hand sides is not trivial.
On the other hand, HPDDM offers support for the
generalized conjugate residual method with inner orthogonalization and deflated
restarting~\cite{parks2006recycling}~(GCRODR), which does not suffer from the aforementioned
limitations.

@@

\pk{KSPHPDDM}, the interface between HPDDM Krylov methods and PETSc, is automatically
registered since PETSc version 3.12 when configuring PETSc with the extra
flag \pk{-{}-download-hpddm}. HPDDM linear solvers can be selected using the command
line option \pk{-ksp\_type hpddm}, or programmatically via \pk{KSPSetType}(\pk{ksp},
\pk{KSPHPDDM}), and the following Krylov methods can be accessed:
\begin{itemize}
    \item pseudo-block GMRES or flexible GMRES~\cite{saad1993flexible};
    \item pseudo-block CG or flexible CG~\cite{notay2000flexible};
    \item pseudo-block GCRODR or flexible GCRODR~\cite{carvalho2011flexible};
    \item block GMRES or flexible GMRES~\cite{calandra2012flexible}, with deflation at each restart;
    \item block CG~\cite{o1980block};
    \item breakdown-free block CG~\cite{ji2017breakdown}, with deflation at each iteration;
    \item block GCRODR or flexible GCRODR, with deflation at each restart.
\end{itemize}
While being mathematically equivalent to
their ``standard'' counterparts, the pseudo-block variants try to fuse together multiple similar
operations like matrix--vector
products to achieve higher arithmetic intensity, or to decrease the
number of global synchronizations needed for scalar products.

HPDDM Krylov methods may be selected using the % {\color{red} TODO API FOR KSPHPDDMSetVariant?}
options \pk{-ksp\_hpddm\_type}
\pk{(gmres$|$cg$|$gcrodr$|$bgmres$|$bcg$|$bfbcg$|$bgcrodr$|$preonly)}, or programmatically 
by using \pk{KSPHPDDMSetType}(\pk{ksp}, \pk{KSPHPDDMType}).
Preconditioning variants can be specified via
\pk{-ksp\_hpddm\_variant} \pk{(left$|$right$|$flexible)} to select whether the
preconditioner is applied on the left, on the right, or if it cannot be
represented as a linear operator. In this latter case, the preconditioner is
always applied on the right, except for the conjugate gradient methods BCG, 
BFCG and BFBCG, that only handle left preconditioning.  It is also
possible to set the preconditioning side through the more common PETSc option
\pk{-ksp\_pc\_side (left$|$right)}.  In addition, convergence monitoring with the
\pk{KSPMonitor} interface is fully supported, as well as the specification of
customized convergence testing via the \pk{KSPSetConvergenceTest} callback.

@@

To fully support HPDDM solvers, from version 3.14.0 of PETSc we added interface routines
for solving systems with multiple right-hand sides, \pk{KSPMatSolve}(\pk{ksp},\pk{X},\pk{Y}),
and to apply preconditioning methods, i.e. \pk{PCMatApply}(\pk{pc},\pk{X},\pk{Y}).
Both input $X$ and output $Y$ matrices are currently limited to be dense tall-and-skinny matrices.
With \pk{KSPHPDDM}, \mbox{(pseudo-)block} methods Krylov methods will be used; instead, when no specialized implementation
is available, PETSc will perform the solution phase
in a column by column fashion.
Furthermore, a function \pk{KSPSetMatSolveBlockSize} is also provided to
decompose a single large block of column vectors into multiple sub-blocks.
\pk{KSPMatSolve} is then called repeatedly until all sub-blocks are traversed.
Similar considerations apply to the underlying calls to \pk{PCMatApply} and \pk{MatProductNumeric}.
This was inspired by MUMPS~\cite{amestoy2001fully} option
ICNTL(27)\footnote{\url{http://mumps.enseeiht.fr/doc/userguide_5.3.3.pdf},
section 6.1}.

While solving the (block) linear system, HPDDM will repeatedly call the following PETSc routines:
\begin{itemize}
    \item \pk{MatMult}($A$, $x$, $y$) for $y = Ax$;
    \item \pk{MatMatMult}($A$, $X$, {\small \pk{MAT\_REUSE\_MATRIX}}, {\small
        \pk{PETSC\_DEFAULT}}, $Y$)\footnote{or \pk{MatProductNumeric}($Y$) with PETSc 3.14.0 and above} for $Y =
        AX$;
    \item \pk{PCApply}(M, $x$, $y$) to apply a preconditioner to $x$, i.e. for $y = M^{-1}x$;
    \item \pk{PCMatApply}(M, $X$, $Y$) for $Y = M^{-1}X$;
\end{itemize}
The rest of the operations are performed directly inside HPDDM.
Specifically, within Krylov methods using the Arnoldi process, or when recycling is requested,
one has to compute $QR$ factorizations of tall-and-skinny dense matrices to orthonormalize
candidate basis vectors. Such operations are performed using the CholQR algorithm~\cite{stathopoulos2002block}, or
via the (modified) Gram--Schmidt method.
For Hessenberg matrices generated by the Arnoldi
process, it is common to update their $QR$ decomposition using Givens
rotations~\cite{saad2003iterative}. For block Hessenberg matrices, Householder reflectors are
used instead~\cite{gutknecht2008updating}.
These orthonormalization variants can be selected at runtime with the option \pk{-ksp\_hpddm\_qr (cholqr$|$mgs$|$cgs)}.

For matrix--matrix products, there are
currently specialized implementations for the following \pk{MatType}:
\begin{itemize}
    \item \pk{MATSEQAIJ}, standard sequential sparse matrix, based on compressed sparse row format;
    \item \pk{MATSEQBAIJ}, block sparse matrix, based on block compressed sparse row format;
    \item \pk{MATSEQSBAIJ}, symmetric block sparse matrix stored in upper triangular form.
    \item \pk{MATMPIAIJ}, standard parallel sparse matrix;
    \item \pk{MATSHELL}, user-defined matrix;
    \item \pk{MATNEST}, block-defined matrix with nested submatrices;
    \item \pk{MATAIJCUSPARSE}, sequential or parallel sparse matrix, offloaded to a GPU using cuSPARSE~\cite{cusparse-web-page}.
\end{itemize}

The following preconditioners have specialized implementations for dealing
efficiently with multiple vectors:
\begin{enumerate}
    %\item \pk{PCNONE}: $P_{A'} = I$, identity matrix;
    \item \pk{PCKSP}: preconditionining via an embedded Krylov;
    \item \pk{PCMAT}: preconditioning via matrix multiplication;
    %\item \pk{PCSHELL}: user-defined preconditioner, e.g., matrix-free;
    % \item \pk{PCSPAI}: sparse approximate inverse~\cite{grote1997parallel};
    \item \pk{PCHARA}~\cite{10.1145/3232850,ambartsumyan2020hierarchical}: hierarchical matrix preconditioners; 
    \item \pk{PCHPDDM}: see~\cref{sec:PCHPDDM};\label{it:hpddm}
    \item \pk{PCASM} and \pk{PCGASM}: overlapping Schwarz methods;\label{it:asm}
    \item \pk{PCBJACOBI}: block Jacobi;\label{it:bjacobi}
    \item \pk{PCLU} or \pk{PCCHOLESKY}: exact $LU$ or Cholesky factorization;\label{it:lu}
    \item \pk{PCILU} or \pk{PCICC}: incomplete $LU$ or Cholesky
        factorization.\label{it:ilu}
\end{enumerate}
One appealing feature of domain decomposition-like preconditioners (i.e. 
\cref{it:hpddm,it:asm,it:bjacobi}) is that they most often rely on exact or inexact
factorizations~ (\cref{it:lu,it:ilu}) as subdomain solvers.
In these cases, it is possible to access the so-called factored matrix $F$
%via \pk{PCFactorGetMatrix} and then call \pk{MatMatSolve}($F$, $\tilde{Y}$, $\tilde{X}$).
via \pk{PCFactorGetMatrix}, and then call \pk{MatMatSolve}($F$, $X$, $Y$) to take advantage
of blocked forward eliminations and backward substitutions from the various factorization packages
interfaced with PETSc. In the case of a sparse matrix,
this strategy is possible with: MUMPS~\cite{amestoy2001fully},
SuiteSparse~\cite{davis2004algorithm}, MKL
PARDISO or CPARDISO~\cite{mkl-web-page}, and SuperLU~\cite{li2005overview} or
SuperLU\_DIST~\cite{li2003superlu_dist}. For the case of a dense
matrix: ScaLAPACK and Elemental~\cite{poulson2013elemental}. 
Future work will consider extending the multigrid framework \pk{PCMG}, as well as other preconditioning classes to increase the arithmetic intensity of the
block preconditioner application phase.

@@

    \item in PETSc with unpreconditioned and unrestarted GMRES($\infty$) \pk{KSPGMRES};

@@

In all the cases, convergence is declared when the initial unpreconditioned residual is reduced by 10 orders of magnitude.

@@

processes and they can be launched using the following command lines: \\[-4pt]

@@

        \caption{\# of iterations needed to converge for various

@@

        the numbers of iterations needed by \pk{KSPHPDDM} match with the respective

@@

In particular, we consider the first three matrix types from~\cref{sec:interface-ksp}, namely \pk{MATSEQAIJ}, \pk{MATSEQBAIJ}, and \pk{MATSEQSBAIJ}.
Their distributed memory counterparts, e.g., \pk{MATMPIAIJ}, rely on the sequential implementations, with each process
storing two such matrices, one with local rows and columns, and another with local rows and ``nonlocal'' columns.
To keep this study succinct, we only evaluate the performance of the sequential implementations, which can help
in drawing conclusions for the intraprocess performance of the parallel formats.
For this reason, from now on, we drop the \pk{SEQ} substring.
The performance of the multiplication primitives of the three aforementioned types will also be compared against
those obtained with the MKL inspector--executor sparse BLAS routines~\cite{mkl-web-page}, using a single OpenMP thread.
There is an ongoing effort to better integrate these routines in PETSc, but at the time of writing, direct calls to the MKL were used.

@@

        xlabel={\# of right-hand sides ($N$)},

@@

    \item with a block size of 1, top plot $b=1$ in~\cref{fig:perfAX}, using any type but \pk{MATAIJ}~\ref{pgfplots:aij1} is counterproductive when $N>1$, see~\ref{pgfplots:baij1} and~\ref{pgfplots:sbaij1}, even when using MKL, see~\ref{pgfplots:mklbaijseq1} and~\ref{pgfplots:mklsbaijseq1}. With 8 or more columns, the performance of \pk{MATAIJ} plateaus and the efficiency reaches approximately 200\%, while the performance of the MKL primitives stagnate at around 175\%, see~\ref{pgfplots:aij1};
for the single column case, \pk{MATSBAIJ} and \pk{MATAIJMKL} deliver slightly better performances;

@@

The CPU configuration which reaches the
highest percentage of peak is \pk{MATSBAIJMKL}~\ref{pgfplots:sbaijmkl3} with $b=3$
and $N=64$, rightmost middle plot in~\cref{fig:perfAXpar}, with approximately
63 GFLOP/s. This format is not yet available in PETSc, but it is handled by the
mini-app. Still, it is less than 1\% of peak for an Intel Xeon Gold 6248.
For the GPU configuration, \pk{MATAIJCUSPARSE}~\ref{pgfplots:cusparse10} with $b=6$
and $N=64$, rightmost bottom plot in~\cref{fig:perfAXpar}, performs at around
306 GFLOP/s, about 4\% of peak for an NVIDIA Tesla V100 SXM2. Future work may
consider using interlaced layouts for storing the dense right-hand sides and extend the
current PETSc functionality in order to maximize performances while maintaining interface flexibility
and the user friendliness of the library.

>>>>

\documentclass[3p,11pt]{elsarticle}

\usepackage{lineno,hyperref}
% \modulolinenumbers[5]

\journal{Computers \& Mathematics with Applications}
\newcommand\col{:}
\newcommand{\pk}[1]{\texttt{#1}}
\usepackage{stmaryrd}
\usepackage{pgfplotstable}
\pgfplotsset{compat=newest}
\pgfplotsset{colormap={paraview}{rgb(0cm)=(0.278431,0.278431,0.858824) rgb(0.1428571429cm)=(0,0,0.360784) rgb(0.2857142857cm)=(0,1,1) rgb(0.4285714286cm)=(0,0.501961,0) rgb(0.5714285714cm)=(1,1,0) rgb(0.7142857143cm)=(1,0.380392,0) rgb(0.8571428571cm)=(0.419608,0,0) rgb(1cm)=(0.878431,0.301961,0.301961)}}
\usepackage{fancyvrb}
\usepackage{subfig}
\usepackage{upquote,textcomp}
\newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
\usepackage{colortbl,booktabs}
\definecolor{last}{RGB}{44,123,182}
\definecolor{blueaccent}{RGB}{215,25,28}
\definecolor{myyellow}{RGB}{102,102,17}
\definecolor{myred}{RGB}{197,65,0}
\definecolor{myblue}{RGB}{0,0,235}
\definecolor{myaqua}{RGB}{9,156,156}
\definecolor{mygreen}{RGB}{63,126,28}
\definecolor{paraviewred}{rgb}{0.278431,0.278431,0.858824}
\definecolor{paraviewblue}{rgb}{0.878431,0.301961,0.301961}
\definecolor{darkgray}{RGB}{253,174,97}
\definecolor{mediumgray}{RGB}{255,255,191}
\usepackage[outline]{contour}
\usepackage{tikz}
\usepackage[binary-units=true]{siunitx}
\usetikzlibrary{decorations.text,arrows.meta,calc,plotmarks,shadows,spy,fixedpointarithmetic,patterns,intersections,arrows.meta,shapes,decorations.text,decorations.pathmorphing,backgrounds,fit,positioning,shapes.symbols,chains,3d,calc,external}
\pgfdeclarelayer{background}
\pgfdeclarelayer{foreground}
\pgfsetlayers{background,main,foreground}
\long\def\ifnodedefined#1#2#3{%
    \@ifundefined{pgf@sh@ns@#1}{#3}{#2}%
}
\makeatother
\usepackage{amsmath,etoolbox,amssymb}
\usepackage{cleveref}
\AtBeginEnvironment{pmatrix}{\setlength{\arraycolsep}{2pt}}
\pgfmathsetmacro{\dof}{4}
\pgfmathsetmacro{\overlap}{1}
\begin{document}

\begin{frontmatter}

\title{KSPHPDDM and PCHPDDM: extending PETS\MakeLowercase{c} with advanced Krylov methods and robust multilevel Overlapping Schwarz preconditioners}

\author[firstaddress]{Pierre Jolivet\corref{correspondingauthor}}
\cortext[correspondingauthor]{Corresponding author}
\ead{pierre.jolivet@enseeiht.fr}

\author[secondaryaddress]{Jose E. Roman}
\ead{jroman@dsic.upv.es}

\author[thirdaddress]{Stefano Zampini}
\ead{stefano.zampini@kaust.edu.sa}

\address[firstaddress]{CNRS, IRIT--ENSEEIHT, Toulouse, France}
\address[secondaryaddress]{Universitat Polit\`ecnica de Val\`encia, Val\`encia, Spain}
\address[thirdaddress]{King Abdullah University of Science and Technology, Thuwal, Saudi Arabia}

% REQUIRED
\begin{abstract}
  Contemporary applications in computational science and engineering often
    require the solution of linear systems which may be of different sizes,
    shapes, and structures. The goal of this paper is to explain how two
    libraries, PETSc and HPDDM, have been interfaced in order to offer
    end-users robust overlapping Schwarz preconditioners and advanced Krylov
    methods featuring recycling and the ability to deal with multiple right-hand sides.
    The flexibility of the implementation %, which for example may deal with various matrix formats,
    is showcased and explained with minimalist, easy-to-run, and reproducible examples,
    to ease the integration of these algorithms into more advanced frameworks.
    The examples provided cover applications from
    %simulate the linear or nonlinear solves of typical
    %applications in fracture mechanics, linear stability analysis or more
    %broadly-speaking
    eigenanalysis, elasticity, combustion, and
    electromagnetism.
\end{abstract}

@@

\end{frontmatter}

\linenumbers

\section{Introduction}
Computational science and engineering today enjoys unprecedented opportunities to transform
the way society solves many of its most urgent technological problems through predictive simulations.
At the heart of simulations are robust and scalable solution algorithms. The efficient
production of applications requires a rich ecosystem of state-of-the-art
reusable libraries from which domain
specialists can benefit~\cite{Bartlett:2017:XFT}. In this paper, we focus on the interoperability of two such
libraries.

@@

Fruits of the proposed enhancement are discussed for applications from
SLEPc~\cite{hernandez2005ssf}, the Scalable Library for Eigenvalue Problem
computations, an add-on library that extends PETSc with classes for
linear and nonlinear eigenvalue problems, as well as with tools to facilitate
the use of shift-and-invert spectral transformations.

The paper is divided in two main parts. \Cref{sec:KSPHPDDM} introduces
the interface to recycling and block Krylov methods
(\pk{KSPHPDDM}) and discusses the readiness of the PETSc library with respect to block Krylov methods.
Different applications from eigenanalysis are also provided.
\Cref{sec:PCHPDDM} introduces the suite of robust multilevel overlapping Schwarz methods (\pk{PCHPDDM})
with examples from linear and nonlinear partial differential equations.
Eventually, concluding remarks are provided in \cref{sec:conclusion}.

@@

Because Krylov methods may need long-term recurrences to mitigate round-off
errors during the generation of subspaces, it is common to introduce a
restart parameter in order to control their memory consumption and the volume of
global communications needed when orthonormalizing a candidate basis vector. Such
restarts may hinder convergence of iterative methods, and even introduce
convergence plateaus. Recycling techniques have been introduced to attenuate
these effects. On the one hand, PETSc implements the loose GMRES~\cite{baker2005technique}~(LGMRES) and the
deflated GMRES~\cite{erhel1996restarted,wakam2013158}~(DGMRES) as
\pk{KSPLGMRES} and \pk{KSPDGMRES}. However, neither handle variable
preconditioning, unlike \pk{KSPFGMRES} or \pk{KSPGCR} which respectively
implement the flexible GMRES~\cite{saad1993flexible}~(FGMRES) and the
generalized conjugate residual method~\cite{eisenstat1983variational}~(GCR).
Furtheremore, \pk{KSPDGMRES} does not support complex arithmetic.  Moreover,
extending the recycling capabilities of such methods to sequence of linear
systems with smoothly varying coefficient matrices and right-hand sides is not
trivial.  On the other hand, HPDDM offers support for the generalized conjugate
residual method with inner orthogonalization and deflated
restarting~\cite{parks2006recycling}~(GCRODR), which does not suffer from the
aforementioned limitations.

@@

\pk{KSPHPDDM}, the interface between HPDDM Krylov methods and PETSc, is
automatically registered since PETSc version 3.12 when configuring PETSc with
the extra flag \pk{-{}-download-hpddm}.  HPDDM linear solvers can be selected
using the command line option \pk{-ksp\_type hpddm}, or programmatically via
\pk{KSPSetType}(\pk{ksp}, \pk{KSPHPDDM}). Then, the following Krylov methods
can be accessed:
\begin{itemize}
    \item pseudo-block GMRES or flexible GMRES~\cite{saad1993flexible};
    \item pseudo-block CG or flexible CG~\cite{notay2000flexible};
    \item pseudo-block GCRODR or flexible GCRODR~\cite{carvalho2011flexible};
    \item block GMRES or flexible GMRES~\cite{calandra2012flexible}, with deflation at each restart;
    \item block CG~\cite{o1980block};
    \item breakdown-free block CG~\cite{ji2017breakdown}, with deflation at each iteration;
    \item block GCRODR or flexible GCRODR, with deflation at each restart.
\end{itemize}
While being mathematically equivalent to
their ``standard'' counterparts, the pseudo-block variants fuse together multiple similar
operations like matrix--vector
products to achieve higher arithmetic intensity, or to decrease the
number of global synchronizations needed for scalar products.

HPDDM Krylov methods may be selected using the % {\color{red} TODO API FOR KSPHPDDMSetVariant?}
options \pk{-ksp\_hpddm\_type}
\pk{(gmres$|$cg$|$gcrodr$|$bgmres$|$bcg$|$bfbcg$|$bgcrodr$|$preonly)}, or programmatically 
by using \pk{KSPHPDDMSetType}(\pk{ksp}, \pk{KSPHPDDMType}).
Preconditioning variants can be specified via
\pk{-ksp\_hpddm\_variant} \pk{(left$|$right$|$flexible)} to select whether the
preconditioner is applied on the left, on the right, or if it cannot be
represented as a linear operator. In this latter case, the preconditioner is
always applied on the right, except for the conjugate gradient methods BCG, 
BFCG, and BFBCG, that only handle left preconditioning.  It is also
possible to set the preconditioning side through the more common PETSc option
\pk{-ksp\_pc\_side (left$|$right)}.  In addition, convergence monitoring with the
\pk{KSPMonitor} interface is fully supported, as well as the specification of
customized convergence testing via the \pk{KSPSetConvergenceTest} callback.

@@

To fully support HPDDM solvers, in version 3.14.0 of PETSc, we added interface routines
for solving systems with multiple right-hand sides, \pk{KSPMatSolve}(\pk{ksp}, \pk{X}, \pk{Y}),
and to apply preconditioners, \pk{PCMatApply}(\pk{pc}, \pk{X}, \pk{Y}).
Both input $X$ and output $Y$ matrices are currently limited to be dense tall-and-skinny matrices.
With \pk{KSPHPDDM}, \mbox{(pseudo-)block} Krylov methods will be used. Instead, when no specialized implementation
is available, PETSc will perform the solution phase
in a column by column fashion.
Furthermore, a function \pk{KSPSetMatSolveBlockSize} is also provided to
decompose a single large block of column vectors into multiple sub-blocks.
\pk{KSPMatSolve} is then called repeatedly until all sub-blocks are traversed.
Similar considerations apply to the underlying calls to \pk{PCMatApply} and \pk{MatProductNumeric}.
This was inspired by MUMPS~\cite{amestoy2001fully} option
ICNTL(27)\footnote{\url{http://mumps.enseeiht.fr/doc/userguide_5.3.3.pdf},
section 6.1}.

While solving linear systems, possibly with multiple right-hand sides, HPDDM will repeatedly call the following PETSc routines:
\begin{itemize}
    \item \pk{MatMult}($A$, $x$, $y$) for $y = Ax$;
    \item \pk{MatMatMult}($A$, $X$, \pk{MAT\_REUSE\_MATRIX}, 
        \pk{PETSC\_DEFAULT}, $Y$)\footnote{or \pk{MatProductNumeric}($Y$) with PETSc 3.14.0 and above} for $Y =
        AX$;
    \item \pk{PCApply}(M, $x$, $y$) to apply a preconditioner to $x$, i.e., for $y = M^{-1}x$;
    \item \pk{PCMatApply}(M, $X$, $Y$) for $Y = M^{-1}X$;
\end{itemize}
The rest of the operations are performed directly inside HPDDM.  Specifically,
within Krylov methods using the Arnoldi process, or when recycling is
requested, one has to compute $QR$ factorizations of tall-and-skinny dense
matrices to orthonormalize candidate basis vectors. Such operations are
performed using the CholQR algorithm~\cite{stathopoulos2002block}, or via the
(modified) Gram--Schmidt method.  These orthonormalization variants can be
selected at runtime with the option \pk{-ksp\_hpddm\_qr (cholqr$|$mgs$|$cgs)}.
For Hessenberg matrices generated by the Arnoldi process, it is common to
update their $QR$ decomposition using Givens
rotations~\cite{saad2003iterative}. For block Hessenberg matrices, Householder
reflectors are used instead~\cite{gutknecht2008updating}.

For matrix--matrix products, there are
currently specialized implementations for the following \pk{MatType}s:
\begin{itemize}
    \item \pk{MATAIJ}: standard sequential or parallel sparse matrix, based on compressed sparse row format;
    \item \pk{MATSEQBAIJ}: block sparse matrix, based on block compressed sparse row format;
    \item \pk{MATSEQSBAIJ}: symmetric block sparse matrix stored in upper triangular form;
    \item \pk{MATSHELL}: user-defined matrix;
    \item \pk{MATNEST}: block-defined matrix with nested submatrices;
    \item \pk{MATAIJCUSPARSE}: sequential or parallel sparse matrix, offloaded to a NVIDIA GPU using cuSPARSE~\cite{cusparse-web-page}.
\end{itemize}
The following preconditioners have specialized implementations for dealing
efficiently with multiple vectors:
\begin{enumerate}
    %\item \pk{PCNONE}: $P_{A'} = I$, identity matrix;
    \item \pk{PCKSP}: embedded Krylov method;
    \item \pk{PCMAT}: matrix multiplication;
    %\item \pk{PCSHELL}: user-defined preconditioner, e.g., matrix-free;
    %\item \pk{PCSPAI}: sparse approximate inverse~\cite{grote1997parallel};
    \item \pk{PCHARA}: hierarchical matrices~\cite{10.1145/3232850,ambartsumyan2020hierarchical}; 
    \item \pk{PCHPDDM}: see~\cref{sec:PCHPDDM};\label{it:hpddm}
    \item \pk{PCASM} and \pk{PCGASM}: overlapping Schwarz methods;\label{it:asm}
    \item \pk{PCBJACOBI}: block Jacobi;\label{it:bjacobi}
    \item \pk{PCLU} or \pk{PCCHOLESKY}: exact $LU$ or Cholesky factorization;\label{it:lu}
    \item \pk{PCILU} or \pk{PCICC}: incomplete $LU$ or Cholesky
        factorization.\label{it:ilu}
\end{enumerate}
One appealing feature of domain decomposition-like
preconditioners~(\cref{it:hpddm,it:asm,it:bjacobi}) is that they most often
rely on exact or inexact factorizations~(\cref{it:lu,it:ilu}) as subdomain
solvers.  In these cases, it is possible to access the so-called factored
matrix $F$
%via \pk{PCFactorGetMatrix} and then call \pk{MatMatSolve}($F$, $\tilde{Y}$,
%$\tilde{X}$).
via \pk{PCFactorGetMatrix}, and then call \pk{MatMatSolve}($F$, $X$, $Y$) to
take advantage of blocked forward eliminations and backward substitutions from
the various factorization packages interfaced with PETSc. In the case of a
sparse matrix, this strategy is possible with: MUMPS~\cite{amestoy2001fully},
SuiteSparse~\cite{davis2004algorithm}, MKL PARDISO or
CPARDISO~\cite{mkl-web-page}, and SuperLU~\cite{li2005overview} or
SuperLU\_DIST~\cite{li2003superlu_dist}. For the case of a dense matrix:
ScaLAPACK and Elemental~\cite{poulson2013elemental} are supported. Future work
will consider extending the multigrid framework \pk{PCMG}, as well as other
preconditioning classes to increase arithmetic intensity of the preconditioner
application phase for blocks of right-hand sides.

@@

    \item in PETSc with no preconditioning and no restart GMRES($\infty$) \pk{KSPGMRES};

@@

In all of the cases, convergence is declared when the initial unpreconditioned
residual is reduced by 10 orders of magnitude.

@@

processes and can be launched using the following command lines. \\[-4pt]

@@

        \caption{Number of iterations needed to converge for various

@@

        the number of iterations needed by \pk{KSPHPDDM} match with the respective

@@

In particular, we consider the first three sequential matrix types from~\cref{sec:interface-ksp}, namely \pk{MATSEQAIJ}, \pk{MATSEQBAIJ}, and \pk{MATSEQSBAIJ}.
Their distributed memory counterparts, e.g., \pk{MATMPIAIJ}, rely on the sequential implementations, with each process
storing two such matrices, one with local rows and columns, and another with local rows and ``nonlocal'' columns.
To keep this study succinct, we only evaluate the performance of the sequential implementations, which can help
in drawing conclusions for the intraprocess performance of the parallel formats.
For this reason, from now on, we drop the \pk{SEQ} substring.
The performance of the multiplication primitives of the three aforementioned types will also be compared against
those obtained with the MKL inspector--executor sparse BLAS routines~\cite{mkl-web-page}, using a single OpenMP thread.
There is an ongoing effort to better integrate these routines in PETSc, but at the time of writing, direct calls to the MKL were used.

@@

        xlabel={Number of right-hand sides ($N$)},

@@

    \item with a block size of 1, top plot $b=1$ in~\cref{fig:perfAX}, using any type but \pk{MATAIJ}~\ref{pgfplots:aij1} is counterproductive when $N>1$, see~\ref{pgfplots:baij1} and~\ref{pgfplots:sbaij1}, even when using the MKL, see~\ref{pgfplots:mklbaijseq1} and~\ref{pgfplots:mklsbaijseq1}. With 8 or more columns, the performance of \pk{MATAIJ} plateaus and the efficiency reaches approximately 200\%, while the performance of the MKL primitives stagnate at around 175\%, see~\ref{pgfplots:aij1}.
For the single column case, \pk{MATSBAIJ} and \pk{MATAIJMKL} deliver slightly better performances;

@@

The CPU configuration which reaches the
highest percentage of peak is \pk{MATSBAIJMKL}~\ref{pgfplots:sbaijmkl3} with $b=3$
and $N=64$, rightmost middle plot in~\cref{fig:perfAXpar}, with approximately
63 GFLOP/s. This format is not yet available in PETSc, but it is handled by the
mini-app. Still, it is less than 1\% of peak for an Intel Xeon Gold 6248.
On the GPU, \pk{MATAIJCUSPARSE}~\ref{pgfplots:cusparse10} with $b=6$
and $N=64$, rightmost bottom plot in~\cref{fig:perfAXpar}, performs at around
306 GFLOP/s, about 4\% of peak for an NVIDIA Tesla V100 SXM2. Future work may
consider using interlaced layouts for storing the dense right-hand sides and extend the
current PETSc functionality in order to maximize performance while maintaining interface flexibility
and the user friendliness of the library.
